---
title: "Traffic in Minneapolis"
output:
  html_document:
    df_print: paged
editor_options:
  markdown:
    wrap: 72
---

# Data Origin
This data originated from UCI's Machine Learning Repository
donated by John Hogue. The data was collected on the busy Interstate 94
between Minneapolis and St. Paul -- a suburb or sibling-city of
Minneapolis.

# Project Goals
The goal of this project was to build a few models to
represent traffic volume on this busy highway, particularly based on the
hours of the day and day of the week. I wanted to get healthy models
built on just this particular set of variables because I was hoping to
isolate the affect of Rain and Cloudiness on weather. In a future
project I hope to incorporate how heavy the rain is and how dense the
clouds are to see if that affects traffic volume more drastically.

```{r setup, include = FALSE}
# Load libraries
# Load any additional libraries. No install.package()
library(lubridate)
library(tidyverse)
library(modelr)

# set.seed for reproducible results
set.seed(1222) 
```

# Load and explore data structure

```{r csv_extract, echo = FALSE}
getwd()
#setwd( #Removed for privacy )
df = read.csv("Metro_Interstate_Traffic_Volume.csv")
glimpse(df)
```

# Data processing

I removed temperatures below Kelvin 230 due to
Minneapolis/St Paul never having recorded temperatures that low. I also
removed any rain values above 1 meter in 1 hour since there was only 1
case of that with a recorded 100,000 millimeters in 1 hour. Obviously,
this would be impossible.

```{r processing, cache = TRUE, dependson = c("setup", "csv_extract")}
traffic = df %>% 
  mutate(
    date_time = as_datetime(date_time)) %>% 
  mutate(
    date = date(date_time),
    wday = wday(date_time, label = TRUE),
    hour = hour(date_time)) %>% 
  filter(
    rain_1h < 1000 & 
    temp > 230) %>%
  mutate(
    sinhour = sin(2 * pi * (hour + 5) / 24))

glimpse(traffic)

traffic_clouds = traffic %>%
  filter(weather_main == "Clouds")
traffic_rain = traffic %>%
  filter(weather_main == "Rain")

```

# Exploration of prediction variables

```{r exploration_charts, message = FALSE}

ggplot(traffic, aes(hour, traffic_volume, color = wday)) +
  geom_smooth() +
  geom_point(alpha = 0.1) +
  labs(title = "First look at Hours vs. Traffic",
       x = "Hours",
       y = "Traffic Volume")

ggplot(traffic, aes(sinhour, traffic_volume, color = wday)) +
  geom_smooth() +
  geom_point(alpha = 0.1) +
  labs(title = "Hours vs. Traffic after using the sin function to linearize hours",
       x = "Sine of Hours",
       y = "Traffic Volume")

ggplot(traffic, aes(x = weather_main, y = traffic_volume, color = weather_main)) +
  geom_boxplot() +
  labs(title = "Weather and Traffic",
       subtitle = "The goal of my project was to see if I could associate cloud cover and rain with less traffic volume",
       x = "Weather Type", 
       y = "Traffic Volume")

```

# Hypothesis Testing

```{r test1}
# Placeholder
t.test(traffic$traffic_volume, traffic_clouds$traffic_volume, alternative = "greater")
t.test(traffic$traffic_volume, traffic_rain$traffic_volume, alternative = "greater")
```

# Split into train / test sets

```{r train_split}
traffic$id = 1:nrow(traffic)
traffic.train = traffic %>% sample_frac(0.70)
traffic.test = traffic %>% anti_join(traffic.train, by = "id")

"If these two values are the same, then the test + train correctly equals the original."
nrow(traffic)
nrow(traffic.train) +
nrow(traffic.test)
```

# Train and test models

## Model 1

Poor R value, but all other metrics are significant. Traffic Volume \~
Hour

```{r model1}

mod1 = lm(traffic_volume ~ hour, data = traffic.train)
summary(mod1)

```

## Model 2

Great R Value, relates Traffic Volume \~ sin(hour) + weekday without
intercept

```{r model2}

mod2 = lm(traffic_volume ~ sinhour + wday - 1, data = traffic.train)
summary(mod2)

```

# Model Performance Comparison

Here we have graphs that are well labeled to display what is going on
within them.

```{r summary, message = FALSE }

traffic.pred = traffic.test %>% 
  spread_residuals(mod1, mod2)

grid = traffic.test %>%
  data_grid(hour = 0:23, wday) %>%
  mutate(sinhour = sin(2 * pi * (hour + 5) / 24)) %>%
  spread_predictions(mod1, mod2) 

ggplot(traffic.pred, aes(hour, traffic_volume)) +
  geom_point(alpha = 0.1) +
  geom_smooth(se = FALSE) +
  geom_smooth(data = grid, aes(y = grid$mod1), color = "red", se = FALSE) +
  geom_smooth(data = grid, aes(y = grid$mod2), color = "hot pink", se = FALSE) +
  labs(title = "Comparing models to reality",
       subtitle = "Blue = original data, Red = Model 1 (linear), Pink = Model 2 (nonlinear)",
       x = "Hour",
       y = "Actual and Predicted Traffic")
  
ggplot(traffic.pred, aes(hour, traffic_volume, color = wday)) +
  geom_point(alpha = 0.1) +
  geom_smooth(se = FALSE, size = 0.2) +
  geom_smooth(data = grid, aes(y = grid$mod1), color = "red", se = FALSE) +
  geom_smooth(data = grid, aes(y = grid$mod2), color = "hot pink", se = FALSE) +
  labs(title = "Comparing models with weekday separation",
       subtitle = "Blue = original data, Red = Model 1 (linear), Pink = Model 2 (nonlinear)",
       x = "Hour",
       y = "Actual and Predicted Traffic") +
  theme(legend.position = "bottom") +
  guides(color = (guide_legend(nrow = 1)))
  

ggplot(traffic.pred, aes(hour, mod1)) +
  geom_smooth(color = "red") +
  geom_smooth(aes(y = mod2), color = "green") +
  geom_point(alpha = 0.1, color = "red") +
  geom_point(aes(y = mod2), alpha = 0.1, color = "green") +
  labs(title = "Residual Plot comparing both models",
       subtitle = "Red (linear), Green (nonlinear)",
       x = "Hour",
       y = "Traffic Volume Residuals")
  
ggplot(traffic, aes(x = weather_main, y = traffic_volume, color = weather_main)) +
  geom_boxplot() +
  labs(title = "Old Weather and Traffic boxes",
       x = "Weather Type", 
       y = "Traffic Volume")

ggplot(traffic.pred, aes(weather_main, mod1, color = weather_main)) + 
  geom_boxplot() +
  labs(title = "New weather boxes based on my linear model",
       x = "Weather",
       y = "Traffic Residuals")
  
ggplot(traffic.pred, aes(weather_main, mod2, color = weather_main)) +
  geom_boxplot() +
  labs(title = "New weather boxes based on my nonlinear model",
       x = "Weather",
       y = "Traffic Residuals")

```

# Final Test

```{r test2}
traffic.pred.clouds = filter(traffic.pred, weather_main == "Clouds")
traffic.pred.rain = filter(traffic.pred, weather_main == "Rain")

t.test(traffic.pred$mod2, traffic.pred.clouds$mod2, alternative = "greater")
t.test(traffic.pred$mod2, traffic.pred.rain$mod2, alternative = "greater")
```

# Analysis and Conclusions

Goal: show that Clouds or Rain has a negative effect on traffic volume
in highway traffic.

Analysis: First I did some data wrangling. Particularly, converting the
date-time which was stored in factor form to more appropriate values and
subsets (i.e. hours and wdays). After this, I tried to identify which
variables likely had the most massive effect on traffic on this
particular highway. I, of course, concluded that the hour of the day is
the foremost value in the data set, then quickly concluded that the wday
would be the second most important. Additionaly, I found that the curve
of hours as an effect on traffic volume looked suspiciously like a
double humped sin wave, so I modified the hours to reflect this possible
distribution. This seemed to make the data more linear, but not
necessarily perfectly. Either way, using these as my baseline for moving
into the models, I worked on making a linear model to decide traffic
volume simply by the hour of the day (univariate). Then, I used the
sin(hours) and weekday as my approace for the multivariate model. This
felt appropriate and it was immediately clear that the multivariate
model was better due to the correlation coefficient being so high.
Finding myself in a decent spot, I started to plot the residuals and was
somewhat satisfied with the distribution. Before all of the modeling I
found that I failed to reject the null hypotheses of rain and clouds
having the same mean as the full data set.

Conclusions: After working with my models and isolating the hour and
weekday variables out, I still wasn't able to show that we could reject
the null hypotheses. So, after all of my hardwork, I found no way to
distinguish between these factors. After doing the project, I feel that
day time and night time would have been another route to isolating the
variable of weather. My hope was that my analysis on hours would have
worked, but I wasn't able to prove anything definitively with the
approach I used to remove hours from the analysis.
